Introduction to R
========================================================
(in just over one hour)
----------------------------------
### February 2014


### 1. Basic R operation
Note: If you are typing into an R script (and not directly to the console), `CTRL+Enter` will submit the current line of code or selection of code.
```{r, highlight=TRUE, tidy=TRUE}
#This is a comment. R ignores this
```

R can do simple arithmetic:
```{r, highlight=TRUE, tidy=TRUE}
2+2 #R ignores everything to hte right of the '#'
```

```{r, highlight=TRUE, tidy=TRUE}
2*3
```

```{r, highlight=TRUE, tidy=TRUE}
2/3
```
  
You can assign data or functions to objects:
```{r, highlight=TRUE, tidy=TRUE}
x <- 2/3
x
```
After creating the object `x`, it should appear in the environment pane in RStudio.
  
In most--but not all--cases, `<-` and `=` are equivalent
```{r, highlight=TRUE, tidy=TRUE}
x = 2/3
x
```
  
Tests of equality use `==` (note that `x` was already assigned a value of 2/3 from the previous step)
```{r, highlight=TRUE, tidy=TRUE}
x == 2/3
```
   
As you would expect, your variables can be used in any other expression
```{r, highlight=TRUE, tidy=TRUE}
x * 6
```

#### Vectors store multiple values
all elements of a vector need to be the same "type" (i.e., numeric or character/factor)
use the `c` function to combine values into an array:
```{r, highlight=TRUE, tidy=TRUE}
y<-c(21,32,23,14,45)
y
```

Accessing parts of a vector:
```{r, highlight=TRUE, tidy=TRUE}
y[1] #note that "1" is for position 1 (unlike some other languages)
y[2:4] # use the colon (:) to refer to a range of elements
```

#### Functions perform tasks
Many functions take a data object as input and perform a calculation. These are simple functions that help you inspect your data objects:
```{r, highlight=TRUE, tidy=TRUE}
#functions can do many things:
length(y)
class(y)
is.numeric(y)
```

#### Missing values
R represents missing values as `NA`, and it is best to explicitly account for missing data.
```{r, highlight=TRUE, tidy=TRUE}
b <- c(1,0,NA)
b == 1
is.na(b)
!(is.na(b))
is.finite(b)
```

### 2. Getting help
Functions also perform (sometimes complex) statistical procedures, and R's built-in help system describes what functions do and how to use them. For example, type `?mean` to see the help page for `mean()`.

A nice feature in RStudio is "tab completion".  If you begin to type the name of a data object or function, pressing tab will open a drop-down menu with all the objects that start with the letters you typed.

You are also encouraged to use Google or other web searches; many people ask and answer questions on the internet, and someone might have answered your question already. Try searching for "Calculate quantiles in Rstats"

### 3. Loading and examining a data set
R can import many types of data formats. Use the correct function to load your data set.
`read.table` and `read.csv` are popular for delimited text files. Otherwise, you might check out the `foreign`, `xlsx`, and `sas7bdat` packages.  In many situations, CSV is preferred.
```{r, highlight=TRUE, tidy=TRUE}
births<-read.csv("http://waisman.wisc.edu/~maenner/wi_birth_2004.csv")
```

You will often have data that is too large to manually inspect. There are several useful functions that will help you examine hte contents of your data set
```{r, highlight=TRUE, tidy=TRUE}
dim(births) #shows number of rows and columns
str(births) #structure of object
summary(births) # summary / overview
colnames(births) #column names of dataframe
head(births) #inspect first several rows
```

#### Accessing parts of a data frame

There are numerous ways to reference parts of a data frame. We're going to focus on bracket notation (e.g., `df[ , ]`) and  `df$` .
```{r, highlight=TRUE, tidy=TRUE}
births[1:5 ,] #first five rows
head(births[,1:5]) # first five columns, `head` function used so all 70000 don't print
```

Referring to variables:
```{r, highlight=TRUE, tidy=TRUE, eval=FALSE}
births$gestAge #try tab complete after the `$`
births[ , 8] #references column number
births[ , 'gestAge'] #references column name in bracket notation
```

Subsetting data:
```{r, highlight=TRUE, tidy=TRUE, eval=FALSE}
boys<-births[births$sex=="M",] # only males
boys<-subset(births, sex== "M") # only males

two_vars<-births[,c("momAge", "smoke")] # 2 columns
two_vars<-subset(births, select=c("momAge","smoke")) # just these 2 columns
```

Let's look at the data again, so we can create a 'clean' dataset
```{r, highlight=TRUE, tidy=TRUE}
summary(births)
table(births$gestAge)
```


Let's subset our data so it only has plausible and complete values for birthweight, maternal smoking, and gestational age
```{r, highlight=TRUE, tidy=TRUE}
births_complete<-births[ births$gestAge <50 &   
                         births$birthweight < 7000 &   
                         births$smoke < 3 &   
                         is.finite(births$smoke),]
```


### 4. Packages and functions
#### Download new packages from CRAN
Getting new packages is a two step process--you first need to download the package from CRAN. (Only the first time you use it, or if you need to update that package).  Before you can use a package's functions in your R session, use the `library()` command to load that package.  Something to note: the package specified in `install.packages()` must be passed as a string, but `library()` will accept strings or non-strings.
```{r, highlight=TRUE, tidy=TRUE, eval=FALSE}
install.packages("deSolve", repos='http://cran.us.r-project.org') #specified repo for knitr
library(deSolve)
```

Function names are followed by parentheses. If you only type the function name, it prints the function code.
```{r, highlight=TRUE, tidy=TRUE, eval=FALSE}
citation  #prints the code
citation() #runs the function
```

Some common functions:
```{r, highlight=TRUE, tidy=TRUE}
table(births$marital, births$smoke)
table(births$marital, births$smoke, useNA="always")
aggregate(x=births_complete$smoke, by=list(births_complete$marital), FUN="mean")
with(births_complete, aggregate(x=smoke, by=list(marital), FUN="mean"))
with(births, chisq.test(table(sex, momRace)))
```

#### regression example
Note that the output that is printed by default does not include the entire object that is generated by the function. 
```{r, highlight=TRUE, tidy=TRUE}
lm(birthweight~gestAge, data=births_complete)
model<-lm(birthweight~gestAge, data=births_complete)
summary(model)
str(model)
confint(model)
plot(model)
```

